Measurement Clusterization in D-optimal Designs for Bayesian Linear Inverse Problems over Hilbert Spaces

### 2023-08-09 latex/prelims.tex
Ready for SIAM applied math?

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces,\footnote{The
subscripts p and o are for ``parameter'' and ``observation'',
respectively.} and let $\fwd: \hilp \to \hilo$ the \emph{forward
operator}. The forward operator $\fwd$ is assumed linear and strongly
smoothing (the heat operator of the 1D heat equation introduced in
Section \ref{section:intro} is a prime example). Take a Gaussian prior
$\param \sim \pr = \normal(\prmean ,\prcov)$ with some appropriate
covariance operator $\prcov$ on $\hilp$ \cite{Stuart10}. Note that
$\fwd \prcov \fwd^*$ is the prior covariance in $\hilo$
\cite{Stuart10}, and as such, $\fwd \prcov \fwd^*$ is assumed
invertible --- an assumption which will be used below\footnote{If
$\fwd$ has a nontrivial kernel we utilize Occam's Razor and ignore
said kernel.}. Measurements are taken via the \emph{measurement
operator}\footnote{It is common for the measurement and forward
operators to be merged $\tmp := \obs \fwd$
\cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
following sections requires that $\fwd$ and $\obs$ are explicitly
separated as in \cite{attia2022stochastic, cvetkovic2023choosing}.}
$\obs \in ( \hilo^* )^m$, where $m$ is the number of measurements
taken. Entries $\meas_j, j=1,\dots,m$ of the observation operator
$\obs$ are called \emph{measurements}:

@@

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error $\eps' \sim
\normal(0,\modcov)$ with $\modcov$ a covariance operator. Observation
error is denoted $\eps \sim \normal(0, \sigma^2 I_m)$, with $I_m \in
\mathbb{R}^{m \times m}$ the identity. Both error terms and the prior
are assumed independent of each other. Thus, data is acquired via
\begin{align}\label{eq:inverse problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}

It is easy to verify that $\obs \eps' + \eps \in \R^m$ is a centered
Gaussian random vector with covariance matrix

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j 
    %
    %
    %
    = \meas_i (\modcov \meas_j).% \text{ (by \eqref{eq:obs*})}.
  \end{split}
\end{align}
Taking $\modcov = 0$ is a common practice
\cite{tarantola2005,Kaipio2005,Vogel02} and then $\Sigma =
\sigma^2I_m$ is a scalar matrix which does not depend on $\obs$.

It is useful to record that the posterior measure $\post$ is Gaussian
in this setting and its covariance operator is \cite{Stuart10}:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}
Note that the poserior covariance $\postcov$ does not depend on data
$\data$.

\subsection{Bayesian D-Optimal Designs in Infinite Dimensions}\label{subsec:D optimal design} 
A Bayesian D-optimal design maximizes expected Kullback-Leibler (KL)
divergence between posterior $\post$ and prior measures $\pr$. It is
useful to first recall the definition of KL divergence \cite{CoverThomas91}:
$$
D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
$$

The study of D-optimal designs for Bayesian linear inverse problems in
infinite dimensions was pioneered by \cite{AlexanderianGloorGhattas14,
  alexanderian2018efficient}. The main result we will make use of is
summarized (in our notation) in Theorem \ref{thm:d optimality}\footnote{In
\cite{AlexanderianGloorGhattas14, alexanderian2018efficient}, results
are stated for $\Sigma=I$ (implied by $\modcov = 0,\sigma^2=1$), but
the authors also mention these results hold for more general
covariance matrices \cite[p. 681]{AlexanderianGloorGhattas14}.}:

\begin{theorem}[Alexanderian, Gloor, Ghattas \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let $\pr = \normal(\prmean,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilp$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}

%% It is important to note that since $\obs$ is finite-rank,
%% $\prcov^{1/2} \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}$ is
%% trace-class.
\begin{definition}\label{def:d optimality}
  $\obs^{\star}$ is said to be \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where entries of $\obs \in (\hilo^*)^m$
  are constrained to some allowed set of measurements in $\hilo^*$.
\end{definition}

For a Bayesian linear model in finite dimensions, with Gaussian prior
and Gaussian noise, a D-optimal design minimizes the determinant of
the posterior covariance matrix\footnote{The frequentist version of
the D-optimality criterion aims to minimize the determinant of the
Fisher information matrix \cite[page 16]{Ucinski05}. Thus, the
Bayesian criterion is just a regularized version of the frequentist
criterion \cite{Chaloner1995}.}. Theorem \ref{thm:d optimality} and
Definition \ref{def:d optimality} carry a similar intuition:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \frac12 \log \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \text{ by \eqref{eq:objective}}\\
    &= \frac12 \log \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \frac12 \log \det \prcov \postcov^{-1} \text{ by \eqref{eq:postcov}}.
    %%  &= \frac12 \log \det \prcov -\frac12 \log \det \postcov.
  \end{split}
\end{align*}
We think of $\prcov$ as constant, so a D-optimal design minimizes the
posterior covariance determinant, analogously to the
finite-dimensional case.

%% \subsection{Notation Summary}\label{subsec:notation}
%%   Our notation is summarized below. We let:
%%   \begin{itemize}
%%   \item $\hilp, \hilo$ Hilbert spaces.
%%   \item $\fwd:\hilp \to \hilo$ a linear compact operator.
%%   \item $\pr \sim \mathcal{N}(0, \prcov)$ prior Gaussian measure on
%%     $\hilp$, with prior covariance operator $\prcov:\hilp \to \hilp$.
%%   \item $\obs: \hilo \to \mathbb{R}^m$ measurement operator, where $m
%%     \in \mathbb{N}$ is the number of measurements taken.
%%   \item $\sigma^2 \in \mathbb{R}_{+}$ observation noise variance.
%%   \item $\modcov$ model error covariance operator.
%%     %% $\data = \obs \fwd \param + \eps$, where $\eps \in
%%     %% \mathbb{R}^m$ isiid $\mathcal{N}(0, \sigma^2)$ noise.
%%   \item $\Sigma(\obs) = \obs \modcov \obs^* + \sigma^2I$. 
%%   \item $\post$ the posterior measure, with covariance $\postcov$.
%%   \item A D-optimality design criterion
%%     \cite{AlexanderianGloorGhattas14}:
%%     \begin{align*}
%%       \begin{split}
%%         \tar(\obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
%%         % 
%%         % 
%%         % 
%%         &= \frac12 \log \det ( I + \prcov^{1/2} \fwd ^* \obs^* \Sigma(\obs)^{-1} \obs
%%         \fwd \prcov^{1/2}).
%%       \end{split}
%%     \end{align*}
%%   %% \item $\{\lambda_i\}_{i=1}^\infty$ eigenvalues of $\fwd\prcov\fwd^*$
%%   %%   in decreasing order of magnitude.
%%   %% %% \item $\{\ev_i\}_{i=1}^\infty$ their corresponding eigenvectors.
%%   %% \item $\{\eta_i\}_{i=1}^\infty$ eigenvalues of $\obs^*\obs$.
%%   \end{itemize}

%% \subsection{Sequential vs Simultaneous Optimization}\label{subsec:seq vs sim}
%% From defintion \ref{def:d optimality} we wish to characterize solution(s) of the
%% following optimization problem for $\tar$. %%: (\hilo^*)^m \to \R$:
%% \begin{align}\label{eq:optimization}
%%   \obs^{\star} := \argmax_{\obs} \tar( \obs ) 
%%   = \argmax_{\obs} \frac12 \log \det 
%%   (I + \prcov^{1/2} \fwd^*\obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}),
%% \end{align}
%% where $\obs$ is constrained to some allowed set of observations. We
%% call this problem ``simultaneous optimization'', since all
%% observations are decided on simulatneously.

%% For computational reasons, one may prefer to find the best
%% observations in a sequential manner. Denote
%% \begin{equation}\label{eq:def obs_k}
%%   \obs_k := (\meas_1,\dots, \meas_k)^t,  k\leq m.
%% \end{equation}
%% Sequential optimal design proceeds as follows. Find $\meas_1$ by
%% maximizing $\tar(\obs_1)$. Then, keeping $\meas_1$ fixed --- find
%% $\meas_2$ as the maximizer of $\tar(\obs_2)$. Then, find $\meas_3$ by
%% keeping $\meas_1,\meas_2$ fixed and taking $\meas_3$ as the maximizer
%% of $\tar(\obs_3)$. Continue this way until $\obs_m = \obs$ is
%% found, where $m$ is the number of available observations. %% It is
%% %% important to notice that this scheme does not require actually
%% %% observing data --- in \eqref{eq:objective} data is averaged out.

>>>>

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let \(\hilp\) and \(\hilo\) be separable Hilbert spaces (the
subscripts p and o are for "parameter" and "observation",
respectively), and let \(\fwd: \hilp \to \hilo\) the \emph{forward
operator}. The forward operator \(\fwd\) is assumed linear and
strongly smoothing (the heat operator of the 1D heat equation
introduced in Section \ref{section:intro} is a prime example). Take a
Gaussian prior \(\param \sim \pr = \normal(\prmean ,\prcov)\) with
some appropriate covariance operator \(\prcov\) on \(\hilp\)
\cite{Stuart10}. Note that \(\fwd \prcov \fwd^*\) is the prior
covariance in \(\hilo\) \cite{Stuart10}, and as such, \(\fwd \prcov
\fwd^*\) is assumed invertible --- an assumption which will be used
below (if \(\fwd\) has a nontrivial kernel we utilize Occam's Razor
and ignore said kernel altogether). Measurements are taken via the
\emph{measurement operator}. It is common for the measurement and
forward operators to be merged \(\tmp := \obs \fwd\)
\cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
following sections requires that \(\fwd\) and \(\obs\) are explicitly
separated as in \cite{attia2022stochastic, cvetkovic2023choosing}.
\(\obs \in ( \hilo^* )^m\), where \(m\) is the number of measurements
taken. Entries \(\meas_j, j=1,\dots,m\) of the observation operator
\(\obs\) are called \emph{measurements}:

@@

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error \(\eps' \sim
\normal(0,\modcov)\) with \(\modcov\) a covariance operator. Observation
error is denoted \(\eps \sim \normal(0, \sigma^2 I_m)\), with \(I_m \in
\mathbb{R}^{m \times m}\) the identity. Both error terms and the prior
are assumed independent of each other. Thus, data is acquired via
\begin{align}\label{eq:inverse problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}

It is easy to verify that \(\obs \eps' + \eps \in \R^m\) is a centered
Gaussian random vector with covariance matrix

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j 
    %
    %
    %
    = \meas_i (\modcov \meas_j).% \text{ (by \eqref{eq:obs*})}.
  \end{split}
\end{align}
Taking \(\modcov = 0\) is a common practice
\cite{tarantola2005,Kaipio2005,Vogel02} and then \(\Sigma =
\sigma^2I_m\) is a scalar matrix which does not depend on \(\obs\).

Finally, it is useful to record that the posterior measure \(\post\) is
Gaussian in this setting and its covariance operator is does not
depend on data \(\data\) \cite{Stuart10}:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

\subsection{Bayesian D-Optimal Designs in Infinite Dimensions}\label{subsec:D optimal design} 
A Bayesian D-optimal design maximizes expected Kullback-Leibler (KL)
divergence between posterior \(\post\) and prior measures \(\pr\). It is
useful to first recall the definition of KL divergence \cite{CoverThomas91}:
\begin{equation*}
  D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
\end{equation*}

The study of D-optimal designs for Bayesian linear inverse problems in
infinite dimensions was pioneered by \cite{AlexanderianGloorGhattas14,
  alexanderian2018efficient}. The main result we will make use of is
summarized in our notation below:

\begin{theorem}[Alexanderian, Gloor, Ghattas \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let \(\pr = \normal(\prmean,\prcov)\) be a Gaussian prior on \(\hilp\)
  and let \(\post = \normal(\postmean,\postcov)\) the posterior measure
  on \(\hilp\) for the Bayesian linear inverse problem \(\data = \obs
  \fwd\param + \obs \eps' + \eps\) discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}

Note that in \cite{AlexanderianGloorGhattas14,
  alexanderian2018efficient}, results are stated for \(\Sigma=I\)
(implied by \(\modcov = 0,\sigma^2=1\)), but the these results also hold
for more general covariance matrices
\cite[p. 681]{AlexanderianGloorGhattas14}.

%% It is important to note that since \(\obs\) is finite-rank,
%% \(\prcov^{1/2} \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}\) is
%% trace-class.
\begin{definition}\label{def:d optimality}
  We say \(\obs^{\star}\) is \emph{D-optimal} if \(\obs^{\star} =
  \argmax_{\obs} \tar(\obs)\), where entries of \(\obs \in (\hilo^*)^m\)
  are constrained to some allowed set of measurements in \(\hilo^*\).
\end{definition}

For a Bayesian linear model in finite dimensions, with Gaussian prior
and Gaussian noise, a D-optimal design minimizes the determinant of
the posterior covariance matrix, and this turns out to be a
regularized version of the frequentist criterion
\cite{Chaloner1995}. Theorem \ref{thm:d optimality} and Definition
\ref{def:d optimality} carry a similar intuition:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \frac12 \log \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \text{ by \eqref{eq:objective}}\\
    &= \frac12 \log \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \frac12 \log \det \prcov \postcov^{-1} \text{ by \eqref{eq:postcov}}.
    %%  &= \frac12 \log \det \prcov -\frac12 \log \det \postcov.
  \end{split}
\end{align*}
We think of \(\prcov\) as constant, so a D-optimal design minimizes a
quantity analogous to the posterior covariance determinant, similarly
to the finite-dimensional case.

%% \subsection{Notation Summary}\label{subsec:notation}
%%   Our notation is summarized below. We let:
%%   \begin{itemize}
%%   \item \(\hilp, \hilo\) Hilbert spaces.
%%   \item \(\fwd:\hilp \to \hilo\) a linear compact operator.
%%   \item \(\pr \sim \mathcal{N}(0, \prcov)\) prior Gaussian measure on
%%     \(\hilp$, with prior covariance operator \(\prcov:\hilp \to \hilp$.
%%   \item \(\obs: \hilo \to \mathbb{R}^m\) measurement operator, where \(m
%%     \in \mathbb{N}\) is the number of measurements taken.
%%   \item \(\sigma^2 \in \mathbb{R}_{+}\) observation noise variance.
%%   \item \(\modcov\) model error covariance operator.
%%     %% \(\data = \obs \fwd \param + \eps$, where \(\eps \in
%%     %% \mathbb{R}^m\) isiid \(\mathcal{N}(0, \sigma^2)\) noise.
%%   \item \(\Sigma(\obs) = \obs \modcov \obs^* + \sigma^2I$. 
%%   \item \(\post\) the posterior measure, with covariance \(\postcov$.
%%   \item A D-optimality design criterion
%%     \cite{AlexanderianGloorGhattas14}:
%%     \begin{align*}
%%       \begin{split}
%%         \tar(\obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
%%         % 
%%         % 
%%         % 
%%         &= \frac12 \log \det ( I + \prcov^{1/2} \fwd ^* \obs^* \Sigma(\obs)^{-1} \obs
%%         \fwd \prcov^{1/2}).
%%       \end{split}
%%     \end{align*}
%%   %% \item \(\{\lambda_i\}_{i=1}^\infty\) eigenvalues of \(\fwd\prcov\fwd^*$
%%   %%   in decreasing order of magnitude.
%%   %% %% \item \(\{\ev_i\}_{i=1}^\infty\) their corresponding eigenvectors.
%%   %% \item \(\{\eta_i\}_{i=1}^\infty\) eigenvalues of \(\obs^*\obs$.
%%   \end{itemize}

%% \subsection{Sequential vs Simultaneous Optimization}\label{subsec:seq vs sim}
%% From defintion \ref{def:d optimality} we wish to characterize solution(s) of the
%% following optimization problem for \(\tar$. %%: (\hilo^*)^m \to \R$:
%% \begin{align}\label{eq:optimization}
%%   \obs^{\star} := \argmax_{\obs} \tar( \obs ) 
%%   = \argmax_{\obs} \frac12 \log \det 
%%   (I + \prcov^{1/2} \fwd^*\obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}),
%% \end{align}
%% where \(\obs\) is constrained to some allowed set of observations. We
%% call this problem ``simultaneous optimization'', since all
%% observations are decided on simulatneously.

%% For computational reasons, one may prefer to find the best
%% observations in a sequential manner. Denote
%% \begin{equation}\label{eq:def obs_k}
%%   \obs_k := (\meas_1,\dots, \meas_k)^t,  k\leq m.
%% \end{equation}
%% Sequential optimal design proceeds as follows. Find \(\meas_1\) by
%% maximizing \(\tar(\obs_1)$. Then, keeping \(\meas_1\) fixed --- find
%% \(\meas_2\) as the maximizer of \(\tar(\obs_2)$. Then, find \(\meas_3\) by
%% keeping \(\meas_1,\meas_2\) fixed and taking \(\meas_3\) as the maximizer
%% of \(\tar(\obs_3)$. Continue this way until \(\obs_m = \obs\) is
%% found, where \(m\) is the number of available observations. %% It is
%% %% important to notice that this scheme does not require actually
%% %% observing data --- in \eqref{eq:objective} data is averaged out.
